---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ximage

<!-- badges: start -->
[![R-CMD-check](https://github.com/hypertidy/ximage/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hypertidy/ximage/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of ximage is to be like 'image()' and 'rasterImage()' but with the
missing functionality and usability problems fixed. Draw an image where you want
it easily

ximage supports making images from

- numeric, character, byte matrix
- numeric array (including RGB/A, and grey/alpha)
- nativeRaster
- data returned from `vapour::gdal_raster_()` or `gdalraster::read_ds()` functions which include numeric, character, byte vectors or nativeRaster types

Missing values display as a colour of your choosing (`na.col`, transparent
by default), a constant `alpha` can be applied over any input for overlay
work, and `zlim` anchors absolute colours across multiple plots.

Alongside `ximage()` there are matching adornment functions that share its
layout conventions: `xcontour()` (contours at cell centres), `xtext()`
(cell values as labels), and `xrect()` (rectangles from extents).

The orientation is in "raster order", i.e. when you're looking at the
picture it's top left to top right, then scan lines down each row to the
bottom (like western reading and `par(mfrow)` order). This orientation is explored in [orientation](https://hypertidy.github.io/ximage/articles/orientation.html) vignette. 

This matches the way that spatial data readers read imagery, and is equivalent
to populating an R matrix with `matrix(<>, nrow, ncol, byrow = TRUE)` (note it's
literally y,x in terms of number of rows then number of columns). Please note
that getting the data out of the matrix is not in this order, but it is with
`as.vector(t(m))`. Higher dimensional arrays need more care, there's no `byrow`
for `array()`. The full story is in the package vignette:
`vignette("orientation", package = "ximage")`.


## Installation

Install the development version from GitHub:

```r
# install.packages("remotes")
remotes::install_github("hypertidy/ximage")
```

## Example

This is a basic example which shows you how to image a topographic dataset,
with arbitrary image overlay placement.

```{r example}
library(ximage)
ximage(topo)  ## plot in the index space of the matrix

## or, plot in the geographic space (we happen to know this for this matrix)
ximage(topo, extent = c(-180, 180, -90, 90), axes = FALSE)
ximage(logo_a, extent = c(170, 180, -40, -30), add = TRUE)

axis(1); axis(2); box()

ximage(logo_a)  ## plot a RGB array, a new plot in its own index space
## overlay a native raster, and a matrix with its own palette
ximage(logo_n, extent = c(10, 20, 20, 40), add = TRUE)
ximage(topo, extent = c(40, 60, 30, 50), add = TRUE, col = hcl.colors(256))
```

Missing values and constant transparency are first-class: here the ocean is
set to missing and displays as `na.col`, and an RGB overlay is applied with
constant `alpha`.

```{r colours}
land <- topo
land[land < 0] <- NA
ximage(topo, extent = c(-180, 180, -90, 90), col = hcl.colors(64, "Blues 3"))
ximage(land, extent = c(-180, 180, -90, 90), add = TRUE,
       col = hcl.colors(64, "Terrain 2"))
ximage(logo_a, extent = c(-60, 60, -80, -35), add = TRUE, alpha = 0.5)
```

We can get imagery from the internet, and plot it very quickly.

```{r imagery}
library(vapour)  
virtual_earth <-  "<GDAL_WMS><Service name=\"VirtualEarth\"><ServerUrl>http://a${server_num}.ortho.tiles.virtualearth.net/tiles/a${quadkey}.jpeg?g=90</ServerUrl></Service><MaxConnections>4</MaxConnections><Cache/></GDAL_WMS>"
par(mar = rep(0, 4))
px <- dev.size("px")
px[which.min(px)] <- 0  ## zero means GDAL fills this dimension, preserving aspect

## change lon_0 and lat_0 to anywhere you like
im <- gdal_raster_nara(virtual_earth, target_ext = c(-1, 1, -1, 1) * 3e5, target_dim = px, target_crs = "+proj=laea +lon_0=147 +lat_0=-42")

system.time(ximage(im, asp = 1))


## crank up the size it's still fast
px <- px * 4
im <- gdal_raster_nara(virtual_earth, target_ext = c(-1, 1, -1, 1) * 3e5, target_dim = px, target_crs = "+proj=laea +lon_0=147 +lat_0=-42")

system.time(ximage(im, asp = 1))

```


## Code of Conduct
  
  Please note that the ximage project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
