---
title: "Why ximage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why ximage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(ximage)
```

## Two functions, and the gap between them

Base R has two ways to draw a grid of pixels, and neither one does the
whole job.

`image()` will map data to colours, but that is where the good news
ends. It cannot handle RGB imagery at all: three channels of colour have
to be heroically de-mapped into a single index plus a constructed
palette before `image()` will touch them. It builds a polygon per cell,
which grinds on grids of any real size. And its `list(x, y, z)` input
convention, with cell-centre coordinate vectors, is bookkeeping that has
nothing to do with the picture.

`rasterImage()` draws colours to the device fast, top-left first, and
has done since 2010. But it *only* takes colours, never data values. It
cannot initiate a plot, something must already have set up a window. It
wants corner coordinates rather than an extent. And it shares no input
structure with `image()`, so there is no smooth path from one to the
other.

So the mapping function cannot draw properly and the drawing function
cannot map, and moving between them means restructuring everything. Then
the icing on the cake: the two disagree about orientation. `image()`
treats a matrix as a field z(x, y), first dimension along x, y
increasing upwards, so the picture is your matrix transposed and flipped
relative to how it prints. `rasterImage()` reads top-left first like
every image format, scanning sensor, and spatial data reader in
existence. This difference is **really confusing**, and it ambushes
everyone eventually.

```{r image-flip}
m <- matrix(as.double(1:12), 3, byrow = TRUE)
m
op <- par(mfrow = c(2, 1), mar = c(2, 2, 2, 1))
image(m, main = "image(): transposed and flipped")
ximage(m, main = "ximage(): what you printed is what you see")
xtext(m, add = TRUE, cex = 2)
par(op)
```

## ximage is the missing middle

ximage does the whole job: it maps values to colours when you give it
data (or passes colours straight through when you already have them,
including RGB arrays and nativeRaster), it initiates the plot itself
from an extent, and it hands the result to `rasterImage()` in one
consistent raster-order convention. That gap between the two base
functions is the motivation for the package, and closing it is
essentially all ximage does.

```{r volcano}
ximage(volcano, extent = c(0, 610, 0, 870))
```

Note: `volcano` as stored is not Maungawhau's real orientation in "north-up" geography, its orientation as a visually striking example for `image()` was established in R's earliest days: see [github volcano package](https://github.com/mdsumner/volcano) for the georeferenced truth.

## Data that already knows its shape

Output from GDAL readers plugs straight in, because those readers report
dimension and extent alongside a flat vector of values in scanline order.
Both the vapour list format and the gdalraster 'gis' attribute format are
understood.

```r
d <- vapour::gdal_raster_data(sds::gebco(), target_dim = c(205, 101))
ximage(d[[1]])   ## a correctly oriented map of the world
```

## Data that doesn't: the dimension guess

A bare vector with no attributes is accepted too, and ximage guesses a
shape by integer-division detection of the length, preferring the most
nearly square factorization and landscape over portrait (in the absence
of information, images are mostly wider than tall). The data is assumed
to be in raster scanline order, exactly the flat output of a GDAL read,
and the message reports the candidate shapes in GDAL dimension order
(ncol x nrow) so you can correct a wrong guess by picking a neighbour
from the list.

```{r guess}
ximage(as.double(1:12))
xtext(matrix(as.double(1:12), 3, byrow = TRUE), add = TRUE, cex = 2)
```

The guess is deliberately conservative: numeric data is always treated as
one plane of values. Only raw vectors are considered for 3-plane (RGB)
and 4-plane (RGBA) pixel-interleaved interpretations, because raw
genuinely is byte data and byte image buffers are overwhelmingly stored
that way.

## Where this points: nativeRaster

The fastest representation of a colour image in R is not a matrix of hex
strings and not a 3-way array of doubles, it is `nativeRaster`: one
packed integer per pixel, each int holding the R, G, B, A bytes, in
raster order. That is the layout `rasterImage()` consumes natively with
no conversion at all, and ximage passes it straight through.

```{r nativeraster}
## a small RGBA image as raw bytes, scanline pixel-interleaved
nc <- 12; nr <- 8
px <- as.raw(c(rbind(seq(0, 255, length.out = nc * nr),   ## R ramp
                     rev(seq(0, 255, length.out = nc * nr)), ## G ramp
                     127,                                   ## B constant
                     255)))                                 ## opaque
## those bytes reinterpret directly as nativeRaster, zero restructuring
nn <- readBin(px, "integer", n = length(px) %/% 4L)
dim(nn) <- c(nr, nc)
class(nn) <- "nativeRaster"
ximage(nn)
```

Notice there is no unpacking step: pixel-interleaved RGBA bytes in
scanline order *are* nativeRaster memory, read four bytes at a time. This
is why the raw plane guess is the one clever case ximage keeps, and why a
3-way array of doubles is usually the wrong destination for colour data
even when it is a correct one. If you are producing colour imagery in R,
produce nativeRaster and everything downstream gets faster.

## Neighbours: png, nara, farver

nativeRaster appears in surprisingly few places across R, and the
packages that do take it seriously each occupy a different corner. The
format itself comes from grDevices (introduced alongside `rasterImage()`
in R 2.11, 2010) but for years the main way to encounter one was
`png::readPNG(native = TRUE)`, and I/O is still the png/jpeg/fastpng
corner: nativeRaster as the thing a decoder hands you and an encoder
accepts, with fastpng adding the width/height/depth attributed raw
vectors that ximage also understands.

coolbutuseless/nara treats nativeRaster as a *canvas*: a mutable
framebuffer of packed ints, drawn on in-place at C speed with points,
lines, polygons, blitting and sprite operations, aimed at realtime
rendering and games. It converts to and from arrays and rasters but its
heart is imperative drawing, and it delegates both I/O and display (the
same `grid.raster()`/`rasterImage()` fast path ximage uses).

farver owns the colour math: vectorized colour space conversion and
encoding without ever touching a string, and `encode_native()` /
`decode_native()` map directly between colour values and the packed
integer format. It is the industrial-strength version of ximage's one
choke point, and if ximage ever took a dependency, farver at `to_hex()`
would be the one worth taking.

ximage's corner is the remaining one: *data* display. Not drawing shapes,
not colour science, not I/O, but the orientation and extent bookkeeping
that connects scientific arrays and flat GDAL reads to `rasterImage()`.
All four packages share the same underlying conviction, that raster
order plus packed pixels plus `rasterImage()` is the fast path R has had
all along, and each takes a different seat around it.

## The shape of the package

ximage follows the hypertidy pattern: a thin wrapper over a fast
primitive, no dependencies beyond base R, and conventions borrowed from
the data formats rather than invented. Everything reduces to one choke
point (values to colours) and one call (`rasterImage()`). The rest is
orientation bookkeeping, which is exactly the part worth getting right
once so nobody has to think about it again.
